A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-12-04 main.tex
Figure 2 boxplot and related edits.

For each $\bX$ matrix, $\bZ$ was a corresponding incidence matrix of 1s and 0s identifying subpopulation, or founding population membership, which assigns the corresponding environmental effect found in $\bgamma$. Various $\bgamma$ patterns were considered in order to investigate the effects of distinct and complex patterns of environmental confounding. Firstly, we considered two different patterns of effect magnitudes, which we refer to as exponential and dichotomous. We use the term dichotomous to describe two effect sizes each experienced by half of the population, regardless of the number of subpopulations, while exponential effects refer to a unique effect size for each subpopulation with exponentially increasing magnitudes. Secondly, we considered whether individuals in genetically similar subpopulations would experience similar environmental effects. We refer to environmental effects that are similar among genetically similar subpopulations as concordant, and to those that are not as discordant.

% For example, a geographic environmental exposure such as the air quality described in our motivating example is likely to be similar among genetically related individuals due to concordance between genetic clusters and geography. However, behavior-based environmental exposures may be less likely to be shared by subpopulations which are genetically and geographically similar. As an example, \citet{ng2014smoking} found considerable within-region variation in smoking rates in Africa, Asia, and South America.

@@

Due to the differing sample sizes and subpopulation structures, a direct comparison of the Empirical data, which has a CP $= 0.87$, with those from the simulated data sets is not straightforward. We note that the CP decreases in the presence of subpopulations of unequal sizes and with more outbred subpopulations as is the case in the Empirical data.

% \anna{MSPE here (?)}

% \anna{talk about karl rohe stuff here - if anywhere}

In addition to MSE, these methods' performance was also assessed in terms of variable selection using the true positive rate (TPR) \anna{technically this is true sign rate - only estimates with the correct sign are counted. Explain this? Calculate overall selection instead?} for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the average TPR for the same settings as in Figure \ref{fig:mse}. The TPRs echo the MSE results in that lasso and LMM-lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the 10 unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-lasso and PC-lasso achieve substantial improvements in TPR over the ordinary lasso. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Empirical data sets than in the 1D Linear Admixture data.

@@

In order to assess the behavior of these methods in this context, we conducted a simulation using the parameter settings presented in Figure \ref{fig:mse} and Table \ref{tab:var_sel} and a modified version of the Empirical data. In order to quantify the absolute difference in subpopulation-specific allele frequencies in an straightforward, interpretable manner, we limited the Empirical data to that of the white and African American subpopulations. Figure \ref{fig:pop_inf} shows the relationship between a given SNP's false positive rate (FPR) and its absolute difference in standardized allele frequency between the two supopulations.

 \begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/pop_inf_snps_loess.png}
    \caption{\anna{Non-causal SNP selection / SNP FPR (?)} in the presence of varying levels of population informative SNPs and environmental confounding for two Empirical subpopulations, dichotomous-discordant effects, $\eta = 0.5$.}
    \label{fig:pop_inf}
\end{figure}

\anna{maybe no table ? If table, make more like table 1}
\begin{table}[H]
\centering
\begin{tabular}{rrlr}
  \hline
eta & xi & Method & $r_s$ \\ 
  \hline
0.5 & 0.0 & lasso & 0.34 \\ 
  0.5 & 0.0 & PC-lasso & -0.92 \\ 
  0.5 & 0.0 & LMM-lasso & -0.89 \\ 
  0.5 & 0.8 & lasso & 0.78 \\ 
  0.5 & 0.8 & PC-lasso & -0.82 \\ 
  0.5 & 0.8 & LMM-lasso & -0.67 \\ 
   \hline
\end{tabular}
\caption{Spearman's rank correlation coefficient between \anna{non-causal SNP selection / SNP FPR (?)} and absolute difference in standardized allele frequencies across methods.}
\label{tab:pop_inf}
\end{table}

\begin{comment}
\anna{Another way to quantify this...(?)}
\begin{table}[H]
\centering
\begin{tabular}{rlrrr}
  \hline
xi & chunks & lasso & lasso\_pca10 & plmm\_lasso \\ 
  \hline
0.0 & 1 & 0.108 & 0.187 & 0.135 \\ 
  0.0 & 2 & 0.394 & 0.132 & 0.221 \\ 
  0.0 & 3 & 0.214 & 0.049 & 0.071 \\ 
  0.8 & 1 & 0.005 & 0.437 & 0.161 \\ 
  0.8 & 2 & 0.504 & 0.188 & 0.348 \\ 
  0.8 & 3 & 0.966 & 0.084 & 0.157 \\ 
   \hline
\end{tabular}
\caption{Max FPR for each method, $\xi$, and level of difference (low, med, high)}
\end{table}
\end{comment}

\anna{Table \ref{tab:pop_inf} displays the Spearman's rank correlation coefficient for each of these scenarios.} In the absence of environmental confounding, the lasso demonstrates a moderately positive correlation with allele frequency differences ($r_s = 0.34$), while the PC-lasso and LMM-lasso methods are negatively correlated with allele frequency differences. Overall, however, the FPR does not exceed 0.39, 0.19, and 0.22 for the lasso, PC-lasso, and LMM-lasso methods, respectively.  
%\anna{We attribute these associations to the presence of population structure (in other words, correlated observations) and thus artificially shrunken errors in the case of the lasso. For PC-lasso and LMM-lasso, the absence of a non-genetic mechanism by which population structure directly affects the outcome, results in a slight over-correction for both, resulting in their moderately negative correlations. }

In the presence of confounding, however, lasso's correlation with allele frequency increases ($r_s = 0.78$), and particularly struggles to correctly select the most population-informative SNPs. PC-lasso and LMM-lasso are able to remain more robust to allele frequency differences, yet remain somewhat negatively correlated with allele frequency differences. This is advantageous in preventing dramatically high FPRs in the presence of highly population-informative SNPs.

In both scenarios, LMM-lasso is more robust to allele frequency differences than PC-lasso due to the presence of \anna{cryptic relatedness in addition to population structure in the Empirical data.} \anna{max absolute kinship? histograms of off-diag kinship?} This echoes the superior TPRs of LMM-lasso compared to that of the PC-lasso in the case of the Empirical data in Table \ref{tab:var_sel}.

>>>>

For each $\bX$ matrix, $\bZ$ was a corresponding incidence matrix of 1s and 0s identifying subpopulation, or founding population membership, which assigns the corresponding environmental effect found in $\bgamma$. Various $\bgamma$ patterns were considered in order to investigate the effects of distinct and complex patterns of environmental confounding. Firstly, we considered two different patterns of effect magnitudes, which we refer to as exponential and dichotomous. We use the term dichotomous to describe two effect sizes each experienced by half of the population, regardless of the number of subpopulations, while exponential effects refer to a unique effect size for each subpopulation with exponentially increasing magnitudes. Secondly, we considered whether individuals in genetically similar subpopulations would experience similar environmental effects. We refer to environmental effects that are similar among genetically similar subpopulations as concordant, and to those that are not as discordant. \anna{We examined all combinations of these patterns of environmental effects, but found they did not result in appreciable differences in performance. We thus focus our attention elsewhere.}

@@

Due to the differing sample sizes and subpopulation structures, a direct comparison of the Empirical data, which has a CP $= 0.87$, with those from the simulated data sets is not straightforward. We note that the CP decreases in the presence of subpopulations of unequal sizes and with more complex population structure as is the case in the Empirical data.

In addition to MSE, these methods' performance was also assessed in terms of variable selection using the true positive rate (TPR) for the first 50 variables each method allowed into the model. \anna{Throughout this work, we refer to a selection as true positive only if it is selected and the sign of the coefficient is estimated correctly.} Table \ref{tab:var_sel} displays the average TPR for the same settings as in Figure \ref{fig:mse}. The TPRs echo the MSE results in that lasso and LMM-lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the 10 unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-lasso and PC-lasso achieve substantial improvements in TPR over the ordinary lasso. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Empirical data sets than in the 1D Linear Admixture data.

@@

In order to assess the behavior of these methods in this context, we conducted a simulation using the parameter settings presented in Figure \ref{fig:mse} and Table \ref{tab:var_sel} and a modified version of the Empirical data. In order to quantify the absolute difference in subpopulation-specific allele frequencies in an straightforward, interpretable manner, we limited the Empirical data to that of the white and African American subpopulations. Figure \ref{fig:pop_inf} summarizes the relationship between a given SNP's false positive rate (FPR) and its absolute difference in MAF between the two supopulations. The top panel of Figure \ref{fig:pop_inf} summarizes the marginal relationship between FPR and MAF differences, while the lower panel shows the association between SNPs falsely selected by each method and MAF differences.

 \begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/pop_inf_snps.png}
    \caption{\anna{Association of population informative nature of SNPs and false selection across methods in the presence and absence of environmental confounding for two Empirical subpopulations, dichotomous-discordant effects, $\eta = 0.5$.}}
    \label{fig:pop_inf}
\end{figure}

%\begin{figure}[H]
%\begin{subfigure}{\textwidth}
%  \centering
%  % include first image
%  \includegraphics[scale = 0.9]{figures/pop_inf_snps_loess.png}  
%  \caption{FPR in the presence of varying levels of population informative SNPs.}
%  \label{fig:pop_inf_scatter}
%\end{subfigure}
%\begin{subfigure}{\textwidth}
%  \centering
%  % include second image
%  \includegraphics[scale = 0.9]{figures/pop_inf_snps_box.png}  
%  \caption{Absolute allele frequency differences of falsely selected, non-causal SNPs across methods.}
%  \label{fig:pop_inf_box}
%\end{subfigure}
%\caption{Association of population informative nature of SNPs and false selection across methods in the presence and absence of environmental confounding for for two Empirical subpopulations, dichotomous-discordant effects, $\eta = 0.5$.}
%\label{fig:pop_inf}
%\end{figure}

\anna{maybe no table ? If table, make more like table 1}
\begin{table}[H]
\centering
\begin{tabular}{rrlr}
  \hline
eta & xi & Method & $r_s$ \\ 
  \hline
0.5 & 0.0 & lasso & 0.42 \\ 
  0.5 & 0.0 & PC-lasso & -0.77 \\ 
  0.5 & 0.0 & LMM-lasso & -0.71 \\ 
  0.5 & 0.8 & lasso & 0.67 \\ 
  0.5 & 0.8 & PC-lasso & -0.61 \\ 
  0.5 & 0.8 & LMM-lasso & -0.34 \\ 
   \hline
\end{tabular}
\caption{\anna{Spearman's rank correlation coefficient between SNP FPR and absolute difference in standardized allele frequencies across methods.}}
\label{tab:pop_inf}
\end{table}

\anna{Table \ref{tab:pop_inf} displays the Spearman's rank correlation coefficient for each of these scenarios.} In the absence of environmental confounding, the lasso demonstrates a moderately positive correlation with allele frequency differences ($r_s = 0.42$), while the PC-lasso and LMM-lasso methods are negatively correlated with allele frequency differences. Overall, however, the FPR does not exceed 0.39, 0.19, and 0.22 for the lasso, PC-lasso, and LMM-lasso methods, respectively.

In the presence of confounding, however, lasso's correlation with allele frequency increases ($r_s = 0.67$), and particularly struggles to correctly select the most population-informative SNPs. PC-lasso and LMM-lasso are able to remain more robust to allele frequency differences, yet remain somewhat negatively correlated with allele frequency differences. This is advantageous in preventing dramatically high FPRs in the presence of highly population-informative SNPs. From a marginal perspective, LMM-lasso seems to exhibit even stronger control of this than PC-lasso. This echoes the superior TPRs of LMM-lasso compared to those of the PC-lasso for the Empirical data in Table \ref{tab:var_sel}. Conditional on selection, however, LMM-lasso may be more likely to confuse MAF differences with true signal.
